\documentclass[11pt,letterpaper]{article}

\author{Jacob Thomas Errington}
\title{Assignment \#4\\Honours set theory -- MATH 488}
\date{10 March 2017}

\usepackage[margin=2.0cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}

\newtheorem{prop}{Proposition}

\newcommand{\inv}{^{-1}}
\newcommand{\Union}{\bigcup}
\newcommand{\Intersn}{\bigcap}
\newcommand{\parens}[1]{\left(#1\right)}
\newcommand{\compl}{\overline}
\newcommand{\intersn}{\cap}

\begin{document}

\begin{prop}
    Let $X$ be a set, $C$ a compact Hausdorff space, and $f : X \to C$ a
    function. SUppose $p$ is an ultrafilter on $X$. Then, there exists a unique
    point $z \in C$ such that for every neighbourhood $U$ of $z$ we have
    \begin{equation*}
        \{ x \in X : f (x) \in U \} \in p
    \end{equation*}
\end{prop}

\begin{proof}
    Suppose not. Then for all $z \in C$, there is a neighbourhood $U_z$ of $z$
    such that $\{ x \in X : f(x) \in U_z \} \notin p$.

    The collection of all such $U_z$ forms a cover of $C$. Hence there exists
    finite $Z_0 \subseteq C$ such that $\{U_z\}_{z \in Z_0}$ is a subcover of
    $C$. Let $A_z = f\inv (U_z)$. Then,
    \begin{equation*}
        \Union_{z \in Z_0} A_z
        = \Union_{z \in Z_0} f\inv (U_z)
        = f\inv \parens{ \Union_{z \in Z_0} U_z }
        = f\inv (C)
        = X
    \end{equation*}
    which shows that $X$ can be finitely partitioned. We take the complement of
    both sides. Since each piece of the partition is not in the ultrafilter,
    each piece's complement is. Hence, the intersection of all the complements,
    which is finite, is also in the ultrafilter, but this is empty set.
    \begin{align*}
        \compl{ \Union_{z \in Z_0} A_z }
        = \Intersn_{z \in Z_0} \compl{A_z}
        = \compl{X}
        = \emptyset
        \in p
    \end{align*}

    This is a contradiction.

    Next, we look at uniqueness. Suppose $z_1, z_2 \in C$ are such that
    $z_1 \neq z_2$. Since $C$ is Hausdorff, we have disjoint neighbourhoods
    $U_{z_1} \ni z_1$ and $U_{z_2} \ni z_2$. By definition of the limit,
    \begin{equation*}
        f\inv (U_{z_1}) \intersn f\inv (U_{z_2}
        = f\inv (U_{z_1} \intersn U_{z_2})
        = \emptyset
        \in p
    \end{equation*}
    which is a contradiction.
\end{proof}

\end{document}
