\documentclass{article}

\usepackage{amsmath, amssymb, amsthm}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{float}
\usepackage[margin=2.5cm]{geometry}

\newtheorem{proposition}{Proposition}

\DeclareMathOperator{\Key}{\mathtt{key}}
\DeclareMathOperator{\Enqueue}{\mathtt{enqueue}}
\DeclareMathOperator{\Dequeue}{\mathtt{dequeue}}
\DeclareMathOperator{\Peek}{\mathtt{peek}}
\DeclareMathOperator{\Remove}{\mathtt{remove}}
\DeclareMathOperator{\Insert}{\mathtt{insert}}
\DeclareMathOperator{\Intersects}{\mathtt{intersects}}
\DeclareMathOperator{\print}{\mathtt{print}}
\DeclareMathOperator{\leftChild}{\mathtt{leftChild}}
\DeclareMathOperator{\rightChild}{\mathtt{rightChild}}
\DeclareMathOperator{\Parent}{\mathtt{parent}}
\DeclareMathOperator{\RemoveMin}{\mathtt{removeMin}}

\author{Jacob Thomas Errington (260636023)}
\title{Assignment \#4\\Honours Algorithms and Data Structures (COMP 252)}
\date{10 March 2015}

\begin{document}

\maketitle

\section{Browsing a red-black tree}

\begin{enumerate}
    \item \emph{Finding $k$-th smallest element.}

        \begin{proposition}
            Given pointers to both the root and the smallest element of a
            red-black tree, finding an element $u$ such that $\Key{[u]} = x$
            can be done in time $O(1 + \log k)$ if $x$ is the $k$-th smallest
            element stored in the tree.
        \end{proposition}

        \begin{proof}
            The sub-trees of a red-black tree are also red-black trees, and so
            are balanced. Between the smallest element and the $k$-th smallest
            element, there are a total of $k$ elements, so the minimal subtree
            that contains these two elements will have height $\log k$. Thus,
            it suffices to find the youngest common ancestor of the two
            elements, since performing the ordinary lookup procedure from there
            will take time $O(\log k)$.

            Starting from the smallest element, we keep going to its parent
            (then its parent's parent, etc.) until we find an element -- call
            it $p$ -- whose \emph{parent's} key is greater than $x$. Therefore,
            $p$ will be the youngest common ancestor of the smallest and the
            $k$-th smallest element. Finding it takes time $O(\log k)$, and
            performing the regular lookup from there takes time $O(\log k)$, so
            overall this algorithm takes time $O(\log k)$.
        \end{proof}

    \item \emph{Listing the $k$ smallest elements.}

        \begin{proposition}
            Given pointers to both the root and the smallest element of a
            red-black tree, the $k$ smallest elements of the tree can be listed
            in time $O(k)$.
        \end{proposition}

        \begin{proof}
            The beginning of this proof is similar. We use the fact that there
            are $k$ elements in the subtree of interest, and we use a modified
            in-order traversal (since it's \emph{bottom-up} rather than
            \emph{top-down} as usual) starting from the smallest element.

            Since the traversal is bottom-up, we split it into two parts: one
            part walks up left leftmost edge of the tree parent by parent, and
            the other part performs augmented in-order traversals of the right
            subtrees of those leftmost elements.

            The traversal is augmented by a counter representing how many
            elements are left to output.  Each time an element is yielded, the
            global counter is decremented, until a total of $k$ elements have
            been yielded.

            \begin{algorithm}[H]
                \caption{List the smallest $k$ elements in a red-black tree
                given the smallest element.}
                \begin{algorithmic}
                    \Require{The smallest element $t$ in a red-black tree, the
                    number of elements to list $k$.}
                    \Ensure{The $k$ smallest elements are printed.}

                    \Function{Traverse}{$t$, $k$}
                        \If{$t$ is \textbf{null} or $k = 0$}
                            \State \Return $k$
                        \Else
                            \State $k^\prime \gets$
                            \Call{Traverse}{$\leftChild{[t]}$, $k$}
                            \State $k^\prime \gets$
                            \Call{Traverse}{$\rightChild{[t]}$, $k^\prime - 1$}
                            \State \Return $k^\prime$
                        \EndIf
                    \EndFunction

                    \Function{ListSmallest}{$t$, $k$}
                        \State $c \gets k - 1$
                        \State $\print{t}$
                        \While{$c > 0$}
                            \State $t \gets \Parent{[t]}$
                            \State $\print{t}$
                            \State $c \gets c - 1$
                            \State $c \gets $
                            \Call{Traverse}{$\rightChild{[t]}$, $c$}
                        \EndWhile
                    \EndFunction
                \end{algorithmic}
            \end{algorithm}

        The algorithm given is $O(k)$ since we essentially perform an in-order
        traversal of the minimal subtree containing the smallest element and
        the $k$-the smallest element, in which there are $k$ elements, all of
        which are printed.
    \end{proof}

\end{enumerate}

\section{Binary heaps}

\begin{proposition}
    The $k$ smallest items of a binary min-heap of size $n$ can be found in
    time $O(k \log k)$.
\end{proposition}

\begin{proof}
    We will create a second min-heap -- call it $H^\prime$ -- of size $k$ in
    which to insert elements as we discover them in the primary heap -- call it
    $H$. We will then traverse $H$ in a fashion prioritized by $H^\prime$, in
    which operations take $O(\log{k})$.

    \begin{algorithm}
        \caption{List the $k$ smallest elements of a binary min-heap.}
        \begin{algorithmic}
            \Require{The number of elements to find $k$, the binary min-heap $H$}
            \Ensure{The $k$ smallest elements are printed.}
            \Function{ListSmallestHeap}{$k$, $H$}
                \State Initialize a min-heap of size $k$, $H^\prime \gets
                \emptyset$
                \State $\Insert{[H^\prime, \Peek{[H]}]}$
                \For{$i = k \to 1$} \Comment{Executes a total of $k$ times}
                    \State $p \gets \RemoveMin{[H]}$ \Comment{$O(\log k)$}
                    \State $\print{p}$
                    \State $\Insert{[H, \leftChild{[p]}]}$
                    \Comment{$O(\log k)$}
                    \State $\Insert{[H, \rightChild{[p]}]}$
                    \Comment{$O(\log k)$}
                \EndFor
            \EndFunction
        \end{algorithmic}
    \end{algorithm}

    Inside the \texttt{for} loop, which is executed $k$ times in total, three
    $O(\log k)$ operations take place, so the overall time of this algorithm is
    $O(3k\log k) = O(k\log k)$ as required.
\end{proof}

\section{Overlapping circles}

\begin{algorithm}
    \caption{Determine if two circles overlap by a sweepline-inspired method}
    \begin{algorithmic}
        \item[]
            The \textsc{CircleOverlap} algorithm uses an interval tree to keep
            track of the $x$-spans of the circles as it sweeps across the
            plane. When two circles are confirmed to overlap on the $x$-axis,
            then they are checked for full overlap.

        \Require{$n$ closed circles in the plane, each determined by a center
        $(x_i, y_i)$ and a radius $r_i$.}

        \Ensure{Whether any two of those circles overlap.}

        \Function{CircleOverlap}{positions $(x_i, y_i)$ and radii $r_i$, for $1 \leq i \leq n$}
            \State Sort the circles by the function $\lambda (x, \_), r \to x - r$.
            \Comment $O(n \log n)$
            %\State Initialize an interval tree $T_x \gets \emptyset$ storing
            %circles, keyed on $\lambda (x, \_), r \to x - r$, with \emph{high}
            %calculated by $\lambda (x, \_), r \to x + r$.
            %\State Initialize an interval tree $T_y \gets \emptyset$ storing
            %circles, keyed on $\lambda (\_, y), r \to y - r$, with \emph{high}
            %calculated by $\lambda (\_, y), r \to y + r$.
            %\State Initialize an empty priority queue $Q \gets \emptyset$ keyed
            %on $\lambda (x, \_), r \to x + r$.
            %\For{$i = 1 \to n$}
            %    \While{$(x_q, y_q), r_q \gets \Peek{[Q]}$} \Comment $O(n \log n)$
            %        \If{$x_q + r_q < x_i - r_i$}
            %            \State $\Remove{[T_x, x_q - r_q]}$
            %            \State $\Remove{[T_y, y_q - y_q]}$
            %            \State $\Dequeue{[Q]}$
            %        \Else
            %            \State \texttt{break}
            %        \EndIf
            %    \EndWhile
            %    \If{$\Intersects{[T_x, ((x_i, y_i), r_i)]}$} \Comment $O(\log n)$
            %        \If{$\Intersects{[T_y, ((x_i, y_i), r_i)]}$} \Comment $O(\log n)$
            %            \State \Return \texttt{True}
            %        \EndIf
            %    \EndIf
            %\EndFor
            \State Initialize an empty interval tree $T_x \gets \emptyset$
            storing circles keyed on $\lambda (x, \_), r \to x - r$, with
            \emph{high} calculated by $\lambda (x, \_), r \to x + r$.
            \State Initialize an empty priority queue $Q \gets \emptyset$
            storing circles keyed on $\lambda (x, \_), r \to x + r$.
            \State $i \gets 2$
            \State $x \gets x_1 - r_1$
            \State $x_{next} \gets x$
            \While{$i \leq n$}
                \State $x_{next} \gets x_i - r_i$
                \If{$x = x_{next}$}
                    \Comment{The sweepline is at the beginning of a new
                        circle.}
                    \If{$cs \gets \Intersects{[T_x, ((x_i, y_i), r_i)]}$}
                        \For{$(x_c, y_c), r_c$ in $cs$}
                            \If{$\sqrt{(x_i - x_c)^2 + (y_i - y_c)^2}
                                    \leq r_i + r_c$}
                                \State \Return \texttt{True}
                            \EndIf
                        \EndFor
                    \EndIf
                    \State $\Enqueue{[Q, ((x_i, y_i), r_i)]}$
                    \State $\Insert{[T_x, ((x_i, y_i), r_i)]}$
                    \State $i \gets i + 1$
                \ElsIf{$(x_q, y_q), r_q \gets \Peek{[Q]}$}
                    \Comment{The sweepline is crossing some circles}
                    \If{$x_q + r_q < x_{next}$} \Comment{The next circle begins
                            after a current one ends.}
                        \State $\Dequeue{[Q]}$
                        \Comment{Remove the circle whose endpoint is soonest.}
                        \State $\Remove{[T_x, ((x_q, y_q), r_q)]}$
                    \EndIf
                \Else
                    \State $x \gets x_{next}$
                    \Comment{Advance the sweepline to the start of the next
                        circle.}
                \EndIf
            \EndWhile
            \State \Return \texttt{False}
        \EndFunction
    \end{algorithmic}
\end{algorithm}

\section{Augmented data structures}

We will use a red-black tree whose nodes will be augmented with two pieces of
information: the difference between their value and their in-order
predecessor's value, and the smallest such difference in the subtree rooted at
that node, i.e. the min-gap in that subtree. The smallest element will be given
the nominal difference $\infty$ with its nonexistent predecessor, and all leaf
nodes will be given the nominal min-gap $\infty$ as they have no subtree.

First, we will remark that the difference between an element and its
predecessor is unaffected by rotations because they preserve order.

Second, we notice that maintaining the min-gap throughout a rotation requires
$O(1)$ time since the group of elements being rotated has at most five
children, collectively, and each of whose min-gap is unchanging. Effectively,
all that will be required during a rotation is to exchange the min-gap of two
of the elements being rotated.

Thus no serious performance penalties are incurred by the self-balancing nature
of the red-black tree when these additional pieces of data are included in the
nodes.

Notice that the min-gap is determined inductively
by the min-gap of a node's two children. Let $p_k$ be
the difference between the value of node $k$ with the value of its predecessor
and $s_k$ be the min-gap of node $k$. Then,
$$s_k = \min\{p_k, p_{\leftChild{[k]}}, s_{\leftChild{[k]}}, p_{\rightChild{[k]}},
s_{\rightChild{:k]}}\}$$.

Now let's examine the running times of the atomic operations.

\begin{description}
    \item[Min-gap.] $O(1)$ since we simply retrieve the min-gap of the root,
        which by the inductive argument made above is the minimal difference
        between any two elements in the tree entire tree.
    \item[Insert.] Upon inserting, which takes time $O(\log n)$, we must locate
        the in-order successor and predecessor of the inserted element, one of
        which takes, at worst, $O(\log n)$ (the other will be $O(1)$ because
        the inserted element is a leaf), and adjust their neighbor-differences
        to account for the new element being added. Due to these changes in
        min-gap, we must travel from parent to parent for both adjusted
        elements, adjusting their min-gap (a kind of \emph{bubble up} process).
        This takes time $O(2 \log n)$. Overall, since only a constant number of
        $O(\log n)$ operations took place, this procedure is $O(\log n)$.
    \item[Delete.] The procedure is essentially the same as for insertion: we
        adjust the neighbor difference of the successor of the deleted element,
        and we bubble up twice to adjust the min-gap. Thus, it operates in
        $O(\log n)$.
    \item[Search.] This procedure is unaffected by the extra data stored in the
        nodes, and so operates in $O(\log n)$.
\end{description}

\end{document}
